Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-06-27 paper/paper.tex
need to figure out arrow placement in figure

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
  Abstracting abstract machines (AAM) is a systematic methodology for constructing 
  abstract interpreters that are derived from concrete small-step abstract machines.
  Recent progress applies the same idea on definitional interpreters, and obtains
  big-step abstract definitional interpreters (ADI) written in monadic style.
  Yet, the relations between small-step abstracting abstract machines and big-step 
  abstracting definitional interpreters are not well studied.

  In this paper, we show their functional correspondence and how to syntactically 
  transform small-step abstract abstract machines into big-step abstract definitional
  interpreters.
  The transformations include linearization, lightweight fusion, disentanglement, 
  refunctionalization, and finally back to direct-style with delimited control operators. 
  Linearization expresses nondeterministic choices by first-order data types, 
  after which refunctionalization transforms the data type representing continuations
  into higher-order functions.
  Following the idea that in deterministic languages, reduction contexts of reduction
  semantics are defunctionalized continuations, we further apply this idea to nondeterministic
  AAM -- its reduction contexts are refunctionalized to extended continuation-passing style.

@@

Defining a language by building an interpreter for it can be traced to the very
early days of programming languages research \cite{landin1966next, Reynolds:72}.
Nowadays, even an undergraduate student in computer science is able to make his own
programming language by building an interpreter.
But building a sound abstract interpreter remained an esoteric and difficult task 
until very recently.

@@

Broadly speaking, abstract abstract machines and abstract definitional interpreters are
different forms of abstract interpreters. They are obtained by applying a combination 
of abstractions to their concrete counterparts, abstract machines and definitional 
interpreters, respectively. An interesting question, and the subject of this paper, is 
how we can interderive these abstract semantic artifacts from the respective other one.

@@

In the domain of abstract semantic artifacts, by contrast, the relations between small-step
abstract abstract machines and big-step abstract definitional interpreters, as well as the 
question of deriving one from the other, are not well studied.
One of the fundamental differences between concrete semantic artifacts and abstract semantic
artifacts is nondeterminism. In addition to ensuring termination, abstract semantic
artifacts are usually equipped with a cache of reachable states, the cache is
updated in a monotonic way and guaranteed to reach the least fixed-point eventually.
In this functional pearl, we make a first step to answering the question of interderiving
abstract semantics artifacts, and provide a constructive answer to relate AAM and ADI 
by presenting a series of syntactical transformations on the program from small-step 
abstract abstract machines to big-step abstract definitional interpreters.

@@

\begin{figure}
\begin{center}
 \begin{tikzpicture}
   \matrix (m) [matrix of math nodes, row sep=2.5em, column sep=3.7em]
     {
         &  &  &  &
       \begin{smallmatrix} \text{abstract} \\ \text{definitional} \\ \text{interpreters} \end{smallmatrix} \\
       \begin{smallmatrix} \text{abstract} \\ \text{abstract} \\ \text{machines} \end{smallmatrix} &
       \begin{smallmatrix} \text{big-step} \\ \text{abstract} \\ \text{abstract} \\ \text{machines} \end{smallmatrix}  &
       \begin{smallmatrix} \text{disentangled}  \\ \text{abstract} \\ \text{abstract} \\ \text{machines} \end{smallmatrix} &
       \begin{smallmatrix} \text{refunctionalized}  \\ \text{abstract} \\ \text{abstract} \\ \text{machines} \end{smallmatrix} &
       \begin{smallmatrix} \text{monadic} \\ \text{abstract} \\ \text{definitional} \\ \text{interpreters} \end{smallmatrix}  \\
       \begin{smallmatrix} \text{abstract} \\ \text{machines} \end{smallmatrix} &
         &
       %\begin{smallmatrix} \text{big-step} \\ \text{abstract} \\ \text{machines} \end{smallmatrix}  &
         &
       %\begin{smallmatrix} \text{disentangled} \\ \text{abstract} \\ \text{machines} \end{smallmatrix} &
         &
       %\begin{smallmatrix} \text{refunctionalized} \\ \text{abstract} \\ \text{machines} \end{smallmatrix} &
       \begin{smallmatrix} \text{definitional} \\ \text{interpreters} \end{smallmatrix} \\
     };
   {
     [start chain] \chainin (m-2-4);
     \chainin (m-1-5) [join={node[sloped,anchor=center,above,labeled] {\text{un-CPS}}}];
     \chainin (m-2-5) [join={node[above,labeled] {\text{}}}];
     \chainin (m-1-5) [join={node[above,labeled] {\text{}}}];
   }
   {
     [start chain] \chainin (m-2-1);
     \chainin (m-2-2) [join={node[above,labeled] {\begin{smallmatrix} \text{linearization} \\ \text{+} \\ \text{fusion} \end{smallmatrix}}     }];
     \chainin (m-2-3) [join={node[above,labeled] {\text{disentang.}}}];
     \chainin (m-2-4) [join={node[above,labeled] {\text{refunc.}}}];
     \chainin (m-2-5) [join={node[above,labeled] {\text{monadify}}}];
   }
   {
     [start chain]
     \chainin (m-3-1);
     { [start branch=A] \chainin (m-2-1)
       [join={node[left, labeled] {\alpha}}];
     }
   }
     %\chainin (m-2-2);
     %\chainin (m-2-3);
     %\chainin (m-2-4);
   {
     [start chain]
     \chainin (m-3-5);
     { [start branch=B] \chainin (m-2-5)
       [join={node[right, labeled] {\alpha}}];
     }
     {
       [start branch=D, color=gray] \chainin (m-3-1)
       [join={node[right, labeled] {\text{}}}];
     }
   }
   {
     [start chain, color=gray] \chainin (m-3-1);
     {
       [start branch=C] \chainin (m-3-5)
       [join={node[above, labeled] {\text{Known as Danvy's correspondences for concrete semantic artifacts }}}];
     }
   }
 \end{tikzpicture}
 \caption{Transformations from AAM to ADI} \label{fig:trans}
\end{center}
\end{figure}

@@

\begin{itemize}
  \item We show \textbf{linearization} in Section~\ref{linear}. By expressing the
    nondeterministic choices as a first-order data type, we linearize the execution
    of abstract abstract machines; the transition of machine states therefore becomes 
    deterministic. Notably, this introduces another layer of control that will also 
    be refunctionalized later.

  \item In Section~\ref{fusing}, we then present the \textbf{lightweight fusion}.
    The fusing transformation simply combines the single-step function @step@ and the 
    driving function into one, but keeps all the machine state representations.

  \item Section~\ref{disen} discusses \textbf{disentanglement}, which disassembles the 
    fused AAM to be several individual functions, with each function handling one data 
    type representing continuation.

  \item \textbf{Refunctionalization} is shown in Section~\ref{sec:refunc} which sequentializes
    the order of abstract execution by higher-order functions. For clarity, we first present 
    the refunctionalized AAM without caching. We then adopt a different caching algorithm to 
    guarantee the termination of abstract interpretation.
    In this section, we also review pushdown control-flow analysis and examine how computable
    and precise call/return matching is obtained through these transformations.

  \item The last transformation we show is that of transforming the refunctionalized AAM to a
    \textbf{direct-style} interpreter (Section~\ref{directstyle}) by using delimited control operators.
\end{itemize}

These transformations are used throughout the paper, with refunctionalization and
defunctionalization of abstract interpreters playing important roles for the call stack of
the analyzed language. By refunctionalization, the call stack of the analyzed language
is blended into the call stack of the defining language. This provides another perspective to
explain why \citeauthor{darais2017abstracting}'s abstract definitional interpreters
is able to inherit the pushdown control-flow property from its defining language.

@@

In this section, we describe a variant of AAM that allows the stack to be
unbounded which uses a precise call stack as we did in the concrete CESK machine.
Instead of allocating continuations in the store and embedding addresses of 
continuations in states, we intend to use a list of frames to explicitly model 
the stack. By doing so, we recover the call stack as the same as runtimes
(so called pushdown control-flow analysis), but since the stack is unbounded, 
the state space is potentially infinite.

@@

In the previous section, we show that by keeping an unabstracted stack in the state
space, we can recover the precise call/return match. Now we begin describing the
transformations step by step. Our base machine for now is the AAM with unbounded
stack, though as we will later transform the stack to higher-order functions
representing continuations, it does not matter what kind of AAM we start from.
This is because transforming to higher-order functions forces us to sequentialize
the order of abstract evaluation, and neither requiring construction of a frame on
stack, nor to allocate continuations in the store.
Thus, our choice to start from an AAM with unbounded stack is motivated simply because
it has an equivalent stack model to abstract definitional interpreters (our final target).

@@

In Danvy's paper \textit{Defunctionalized Interpreters for Programming Languages},
he mentions that for deterministic languages, a reduction semantics is a structural
operational semantics in continuation-passing style, where the reduction context is
a defunctionalized continuation \cite{Danvy:2008:DIP:1411204.1411206}. Therefore
refunctionalizing a deterministic language yields an interpreter in continuation-passing
style; nevertheless, doing so for nondeterministic languages yields an interpreter
with multiple layers of continuation \cite{Danvy:2006:RW:2171265.2171268}.

@@

Refunctionalization transforms first-order data types representing
evaluation contexts to higher-order functions.
Besides sequentializing the order of abstract evaluation by functions,
this transformation can be also regarded as allocating continuation functions of
the defined language in the heap of the metalanguage.

@@

\textit{The Problem with Return-flows.}
Pushdown control-flow is a property in analysis that precisely models the runtime
call stack of the analyzed program. A pushdown control-flow analysis provides an 
as-exact-as-runtime return-flow when analyzing a program, but traditional 
control-flow analysis collapses the state space into a finite space and thus 
causes imprecise stack modeling.

@@

In fact, what we started is an AAM with an unbounded stack which already precisely
matches the calls and returns. Refunctionalization makes the interpreter be a ``big-step''
semantic artifact, then the consequence is we have no places to explicitly save the context
information. But refunctionalization not only forces us to sequentialize the order of
nondeterministic evaluation through higher-order functions representing continuations,
but also drives us to use a new caching algorithm.

@@

Over the years, Danvy and his collaborators applied refunctionalization and defunctionalization 
to many different abstract machines and interpreters and build functional correspondences among
them \cite{Ager:2003:FCE:888251.888254, Danvy:2001:DW:773184.773202, danvy2004refocusing,
Danvy:2008:DIP:1411204.1411206, AGER2004223, ager2005functional, Danvy:2006:RW:2171265.2171268,
danvy2009towards, biernacka2009towards}. Such semantic artifacts include the CEK machines,
the CLS machines, the SECD machines, call-by-need evaluators, monadic evaluators, etc.
Other applications are also found to be applicable, such as Dyck word recognizer, and 
Dijkstra's shunting-yard algorithm and even Quicksort algorithm \cite{Danvy:2006:RW:2171265.2171268}.
For deterministic languages, \citeauthor{Danvy:2008:DIP:1411204.1411206} shows a reduction semantics
is a structural operational semantics in continuation-passing style, where the reduction contexts
are defunctionalized (data type representing) continuations \cite{Danvy:2008:DIP:1411204.1411206}.
Not unexpectedly, the correspondence also exists for nondeterministic languages. For example,
\citeauthor{Danvy:2001:DW:773184.773202} show that refunctionalizing a regualr expressions matcher yields
its counterpart with two layer of continuations \cite{Danvy:2001:DW:773184.773202}.
And the multi-layer continuation-passing style is the underlying semantics of control operatos
like @reset@ and @shift@ \cite{Danvy:1990:AC:91556.91622}.

@@

Lightweight fusion as one of the transformation in our paper is also studied by 
\citeauthor{Ohori:2007:LFF:1190216.1190241} \cite{Ohori:2007:LFF:1190216.1190241}. \\

@@

\textit{Abstracting Definitional Interpreters.}
Reynold's seminal paper \emph{Definitional interpreters for higher-order programming languages} 
\cite{Reynolds:72, Reynolds:HOSC98-revisited} shows that in some definitional interpreters, the
defined language can inherit properties from the defining language. With this insight,
\citeauthor{darais2017abstracting} construct abstracting definitional interpreters which
automatically inherit the pushdown control-flow property from its defining language because
the defined language simply uses the call stack of the meta-language~\cite{darais2017abstracting}.
\citeauthor{darais2017abstracting}'s abstracting definitional interpreters work on direct-style
$\lambda$-calculus, and itself is written in monadic style. One of the advantages of monadic
abstract interpreters is modular and composable, therefore deploying different sensitivities
or features is just applying a different monad. Prior to that, \citeauthor{Sergey:2013:MAI:2491956.2491979}
show a monadic abstract interpreter for small-step semantics \cite{Sergey:2013:MAI:2491956.2491979}.

This paper is greatly inspired by~\citeauthor{darais2017abstracting}'s work. The difference on the
surface is that ours works for A-Normal Form $\lambda$-calculus, instead of plain $\lambda$-calculus,
although our work can also be easily adapted to handle a plain $\lambda$-calculus. But besides directly
presenting the final form of abstract definitional interpreters, we reveal that refunctionalization
plays an important role for inheriting the call stack from the meta-language. The target of the series
of transformations in this paper is abstract definitional interpreters, and we additionally demonstrate
abstract interpreters written direct-style with delimited control operators. We use the delimited control
operators @shift@ and @reset@ that are implemented in Scala \cite{rompf2009implementing}. The
correspondence of delimited control (as well as continuations) and monads are well-known
\cite{Danvy:1990:AC:91556.91622, wadler1992essence, danvy1992representing, moggi1991notions}.

@@

In this functional pearl, we fill the gap between small-step abstract abstract machines 
and big-step abstract definitional interpreters by applying a series of syntactical 
transformations. Among these transformations, linearization turns a worklist into another
layer of continuations; refunctionalization converts the first-order data types representing
continuations to higher-order functions; and finally un-CPS with delimited control transforms
the abstract interpreter into a sequence of expressions, which looks no different to a
concrete interpreter.

We show the functional correspondence not only exists between concrete semantic artifacts
but also exists between abstract semantic artifacts. An interesting open question would
be whether there also exists a correspondence of static analyses formalized in different 
denotational style and operational style.

>>>>

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
  Abstracting abstract machines (AAM) is a systematic methodology for constructing 
  abstract interpreters that are derived from concrete small-step abstract machines.
  \citeauthor{darais2017abstracting} apply the same idea on monadic definitional interpreters, 
  and obtain monadic abstract definitional interpreters (ADI) in big-step style.
  Yet, the relations between small-step abstracting abstract machines and big-step 
  abstracting definitional interpreters are not well studied.

  In this paper, we show their functional correspondence and how to syntactically 
  transform small-step abstract abstract machines into big-step abstract definitional
  interpreters.
  The transformations include linearization, lightweight fusion, disentanglement, 
  refunctionalization, and finally back to direct-style with delimited control operators. 
  Linearization expresses nondeterministic choices by first-order data types, 
  after which refunctionalization transforms the first-order datatypes representing continuations
  into higher-order functions. 
  The refunctionalized AAM is an abstract interpreter written in continuation-passing style
  with two layers of continuation.

  %Following the idea that in deterministic languages, reduction contexts of reduction
  %semantics are defunctionalized continuations, we further apply this idea to nondeterministic
  %AAM -- its reduction contexts are refunctionalized to continuation-passing style.

@@

Defining a language by building an interpreter for it can be traced to the very
early days of programming languages research \cite{landin1966next, Reynolds:72, 
McCarthy:1960:RFS:367177.367199}.
Nowadays, even undergraduate students in computer science are able to make their own
programming language by building an interpreter.
But building a sound abstract interpreter remained an esoteric and difficult task 
until very recently.

@@

Broadly speaking, abstract abstract machines and abstract definitional interpreters are
different forms of abstract interpreters. They are obtained by applying a combination 
of abstractions to their concrete counterparts, abstract machines and definitional 
interpreters, respectively. An interesting question, and the subject of this paper, is 
how we can derive an abstract semantic artifact from the respective other one.

@@

In the domain of abstract semantic artifacts, by contrast, the relations between small-step
abstract abstract machines and big-step abstract definitional interpreters, as well as the 
question of deriving one from the other, are not well studied.
One of the fundamental differences between concrete semantic artifacts and abstract semantic
artifacts is a sound latter one must tell you facts about all possible execution paths, and 
thus nondeterminism is introduced. In addition to ensuring termination, abstract semantic
artifacts are usually equipped with a cache of reachable states, the cache is
updated in a monotonic way and guaranteed to reach the least fixed-point eventually.
In this functional pearl, we make a first step to answering the question of interderiving
abstract semantics artifacts, and provide a constructive answer to relate AAM and ADI
by presenting a series of syntactical transformations on the program from small-step 
abstract abstract machines to big-step abstract definitional interpreters.

@@

\begin{figure}
\begin{center}
 \begin{tikzpicture}
   \matrix (m) [matrix of math nodes, row sep=2.5em, column sep=3.7em]
     {
         &  &  &  &
       \begin{smallmatrix} \text{abstract} \\ \text{definitional} \\ \text{interpreters} \end{smallmatrix} \\
       \begin{smallmatrix} \text{abstract} \\ \text{abstract} \\ \text{machines} \end{smallmatrix} &
       \begin{smallmatrix} \text{big-step} \\ \text{abstract} \\ \text{abstract} \\ \text{machines} \end{smallmatrix}  &
       \begin{smallmatrix} \text{disentangled}  \\ \text{abstract} \\ \text{abstract} \\ \text{machines} \end{smallmatrix} &
       \begin{smallmatrix} \text{refunctionalized}  \\ \text{abstract} \\ \text{abstract} \\ \text{machines} \end{smallmatrix} &
       \begin{smallmatrix} \text{monadic} \\ \text{abstract} \\ \text{definitional} \\ \text{interpreters} \end{smallmatrix}  \\
       \begin{smallmatrix} \text{abstract} \\ \text{machines} \end{smallmatrix} &
         &
       %\begin{smallmatrix} \text{big-step} \\ \text{abstract} \\ \text{machines} \end{smallmatrix}  &
         &
       %\begin{smallmatrix} \text{disentangled} \\ \text{abstract} \\ \text{machines} \end{smallmatrix} &
         &
       %\begin{smallmatrix} \text{refunctionalized} \\ \text{abstract} \\ \text{machines} \end{smallmatrix} &
       \begin{smallmatrix} \text{definitional} \\ \text{interpreters} \end{smallmatrix} \\
     };
   {
     [start chain] \chainin (m-2-4);
     \chainin (m-1-5) [double,join={node[sloped,anchor=center,above,labeled] {\text{CPS left-inv.}}}];
     \chainin (m-2-5) [join={node[above,labeled] {\text{}}}];
     \chainin (m-1-5) [join={node[above,labeled] {\text{}}}];
   }
   %{ [start chain] \chainin (m-1-5);
   % \chainin (m-2-4) [join={node[sloped,anchor=center,above,labeled] {\text{CPS trans.}}}]; }
   %
   % TODO: find out how to make the arrows visually separate
   % {
   %   [start chain] \chainin (m-1-5);
   %   \chainin (m-2-4) [bend left=5, join={node[sloped,anchor=center,below,labeled] {\text{CPS}}}];
   % }
   {
     [start chain] \chainin (m-2-1);
     \chainin (m-2-2) [join={node[above,labeled] {\begin{smallmatrix} \text{linearization} \\ \text{+} \\ \text{fusion} \end{smallmatrix}}     }];
     \chainin (m-2-3) [join={node[above,labeled] {\text{disentang.}}}];
     \chainin (m-2-4) [join={node[above,labeled] {\text{refunc.}}}];
     %\chainin (m-2-5) [join={node[above,labeled] {\text{monadify}}}];
   }
   {
     [start chain, color=gray]
     \chainin (m-3-1);
     { [start branch=A] \chainin (m-2-1)
       [join={node[left, labeled] {\alpha}}];
     }
   }
     %\chainin (m-2-2);
     %\chainin (m-2-3);
     %\chainin (m-2-4);
   {
     [start chain, color=gray]
     \chainin (m-3-5);
     { [start branch=B] \chainin (m-2-5)
       [join={node[right, labeled] {\alpha}}];
     }
     {
       [start branch=D, color=gray] \chainin (m-3-1)
       [join={node[right, labeled] {\text{}}}];
     }
   }
   {
     [start chain, color=gray] \chainin (m-3-1);
     {
       [start branch=C] \chainin (m-3-5)
       [join={node[above, labeled] {\text{Known as Danvy's correspondences for concrete semantic artifacts }}}];
     }
   }
 \end{tikzpicture}
 \caption{Transformations from AAM to ADI} \label{fig:trans}
\end{center}
\end{figure}

@@

\begin{itemize}
  \item We show \textbf{linearization} in Section~\ref{linear}. By expressing the
    nondeterministic choices as a first-order data type, we linearize the execution
    of abstract abstract machines; the transition of machine states therefore becomes 
    deterministic. Notably, this explicitly expresses another layer of control that 
    will also be refunctionalized later.

  \item In Section~\ref{fusing}, we then apply a \textbf{lightweight fusion} 
    \cite{DANVY2008100, Ohori:2007:LFF:1190216.1190241} to the linearized AAM. 
    The fusing transformation simply combines the single-step function @step@ and the 
    driving function into one, but keeps all the machine state representations.

  \item Section~\ref{disen} discusses \textbf{disentanglement}. In this part, we identify 
    different first-order datatypes which represent different layers of continuation, and
    disassemble their handlers to be separate higher-order functions.

  \item Applying \textbf{refunctionalization} \cite{DANVY2009534, Danvy:2006:RW:2171265.2171268} 
    to disentangled AAM is shown in Section~\ref{sec:refunc}. 
    %as the left inverse of Reynolds' defunctionalization \cite{Reynold:72,Reynolds:HOSC98-revisited} 
    Refunctionalization replaces the first-order datatypes representing continuations with
    higher-order functions, and associates dispatching logic with proper higher-order functions. 
    We then derive an abstract interpreter written in continuation-passing style with two
    layers of continuations due to the nondeterminism.

    For clarity, we first present the refunctionalized AAM without caching. 
    We then adopt a caching algorithm \cite{darais2017abstracting} to guarantee the termination 
    of abstract interpretation.
    In the end of this section, we review the pushdown control-flow analysis and examine how 
    computable and precise call/return matching is obtained through these transformations.

  \item The last transformation we show is that of transforming the refunctionalized AAM back to a
    \textbf{direct-style} interpreter (Section~\ref{directstyle}) by using delimited control operators.
\end{itemize}

These transformations are used throughout the paper, with refunctionalization and
defunctionalization of abstract interpreters playing important roles for the stack model of
the analyzed language. By refunctionalization, the call structure of the analyzed language
is blended into the call structure of the defining language. This provides another perspective to
explain why \citeauthor{darais2017abstracting}'s abstract definitional interpreters
is able to inherit the pushdown control-flow property from its defining language.

@@

In this section, we describe a variant of AAM that allows the stack to be unbounded 
which uses a precise model of call stack as we did in the concrete CESK machine.
Instead of allocating continuations in the store and embedding addresses of 
continuations in states, we intend to use a list of frames to explicitly model 
the stack. By doing so, we recover the call structure as the same as runtimes
(so called pushdown control-flow analysis), but since the stack is unbounded, 
the state space is potentially infinite.

@@

In the previous section, we show that by keeping an unabstracted stack in the state
space, we can recover the precise call/return match. Now we begin describing the
transformations step by step. Our base machine for now is the AAM with unbounded
stack, though as we will later transform the stack structure to higher-order functions
representing continuations, it does not matter what kind of AAM we start from.
This is because it would not be possible to construct a new frame on the stack, nor
to allocate continuations in the store.
Thus, our choice to start from an AAM with unbounded stack is motivated simply because
it has an equivalent stack model to abstract definitional interpreters (our final target).

@@

In Danvy's paper \textit{Defunctionalized Interpreters for Programming Languages},
he mentions that for deterministic languages, a reduction semantics is a structural
operational semantics in continuation-passing style, where the reduction context is
a defunctionalized continuation \cite{Danvy:2008:DIP:1411204.1411206}. Therefore
refunctionalizing a deterministic language yields an interpreter in continuation-passing
style. Nevertheless, doing so for nondeterministic languages yields an interpreter
with multiple layers of continuation \cite{Danvy:2006:RW:2171265.2171268, DANVY2009534},
where the additional layer of continuation represents the nondeterministic choices.

@@

Refunctionalization transforms first-order data types representing evaluation contexts 
to higher-order functions, and associates the dispatching logic of continuations with
proper higher-order functions.
This transformation can be also regarded as making the control flows of the defined 
language be the part of control flows of the meta-language.
%allocating continuation functions of the defined language in the heap of the meta-language.

@@

\textit{The Problem with Return-flows.}
Pushdown control-flow is a property in analysis that precisely models the runtime
call structure of the analyzed program. A pushdown control-flow analysis provides an 
as-exact-as-runtime return-flow when analyzing a program, but traditional 
control-flow analysis collapses the state space into a finite space and thus 
causes imprecise stack modeling.

@@

In fact, what we started is an AAM with an unbounded stack which already precisely
matches the calls and returns. Refunctionalization makes the interpreter be a ``big-step''
semantic artifact, then the consequence is we have no places to explicitly save the context
information in store or stack. 
But refunctionalization not only forces the nondeterministic control flow to be higher-order,
but also drives us to use a different caching algorithm.

@@

Over the years, Danvy and his collaborators applied refunctionalization and defunctionalization 
to many different abstract machines and interpreters and build functional correspondences among
them \cite{Ager:2003:FCE:888251.888254, Danvy:2001:DW:773184.773202, danvy2004refocusing,
Danvy:2008:DIP:1411204.1411206, AGER2004223, ager2005functional, Danvy:2006:RW:2171265.2171268,
DANVY2009534, danvy2009towards, biernacka2009towards}. Such semantic artifacts include the CEK machines,
the CLS machines, the SECD machines, call-by-need evaluators, monadic evaluators, etc.
Other applications are also found to be applicable, such as Dyck word recognizer, and 
Dijkstra's shunting-yard algorithm and even Quicksort algorithm \cite{Danvy:2006:RW:2171265.2171268,
DANVY2009534}.
For deterministic languages, \citeauthor{Danvy:2008:DIP:1411204.1411206} shows a reduction semantics
is a structural operational semantics in continuation-passing style, where the reduction contexts
are defunctionalized (data type representing) continuations \cite{Danvy:2008:DIP:1411204.1411206}.
Not unexpectedly, the correspondence also exists for nondeterministic languages. For example,
\citeauthor{Danvy:2001:DW:773184.773202} show that refunctionalizing a regualr expressions matcher yields
its counterpart with two layer of continuations \cite{Danvy:2001:DW:773184.773202}.
And the multi-layer continuation-passing style is the underlying semantics of control operatos
like @reset@ and @shift@ \cite{Danvy:1990:AC:91556.91622}.

@@

Lightweight fusion as one of the transformation in our paper is also studied by 
\citeauthor{Ohori:2007:LFF:1190216.1190241} \cite{Ohori:2007:LFF:1190216.1190241}.
As shown by \citeauthor{DANVY2008100}, lightweight fusion can be used to show the 
equivalence between small-step and big-step abstract machines \cite{DANVY2008100}.\\

@@

\textit{Abstracting Definitional Interpreters.}
Reynold's seminal paper \emph{Definitional interpreters for higher-order programming languages} 
\cite{Reynolds:72, Reynolds:HOSC98-revisited} shows that in some definitional interpreters, the
defined language can inherit properties from the defining language. With this insight,
\citeauthor{darais2017abstracting} construct abstracting definitional interpreters which
automatically inherit the pushdown control-flow property from its defining language because
the defined language simply uses the call-stack model of the meta-language~\cite{darais2017abstracting}.
\citeauthor{darais2017abstracting}'s abstracting definitional interpreters work on direct-style
$\lambda$-calculus, and itself is written in monadic style. One of the advantages of monadic
abstract interpreters is modular and composable, therefore deploying different sensitivities
or features is just applying a different monad. Prior to that, \citeauthor{Sergey:2013:MAI:2491956.2491979}
show a monadic abstract interpreter for small-step semantics \cite{Sergey:2013:MAI:2491956.2491979}.

This paper is greatly inspired by~\citeauthor{darais2017abstracting}'s work \cite{darais2017abstracting}.
The difference on the surface is that ours works for A-Normal Form $\lambda$-calculus, instead of plain 
$\lambda$-calculus, although our work can also be easily adapted to handle a plain $\lambda$-calculus. 
But besides directly presenting the final form of abstract definitional interpreters, we reveal that 
refunctionalization plays an important role for inheriting the model of the stack from the meta-language. 
The target of these transformations in this paper is abstract definitional interpreters, and we 
additionally demonstrate abstract interpreters written direct-style with delimited control operators. 
We use the delimited control operators @shift@ and @reset@ that are implemented in Scala 
\cite{rompf2009implementing}. The correspondence of delimited continuations and monads are well-known
\cite{Danvy:1990:AC:91556.91622, wadler1992essence, danvy1992representing, moggi1991notions}.

@@

In this functional pearl, we fill the gap between small-step abstract abstract machines
and big-step abstract definitional interpreters by applying a series of syntactical
transformations. Among these transformations, linearization turns a worklist into another
layer of continuations; refunctionalization converts the first-order data types representing
continuations to higher-order functions; and finally a left-inverse CPS transformation turns
the abstract interpreter into a sequence of expressions with delimited control operators,
which looks almost no different to a concrete interpreter.

We show the functional correspondence not only exists between concrete semantic artifacts
but also exists between abstract semantic artifacts. An interesting open question would
be whether there also exists correspondences between static analyses that are formalized 
in different denotational and operational styles.
